Unifying the communicable disease spreading paradigm with Gompertzian growth

### 2022-07-30 ScientificReports/main.tex
add sciRep manuscript

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{float}
\usepackage{booktabs}       % professional-quality tables
\usepackage{graphicx}
\modulolinenumbers[5]
\journal{Chaos, Solitons \& Fractals}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
% \bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
% \bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\insertPdfFig}[3]{
  \begin{figure}[H]
  \centering
  \includegraphics[width=12.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}

\newcommand{\insertSmallPdfFig}[3]{
  \begin{figure}[h]
  \centering
  \includegraphics[width=8.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}
\graphicspath{{../figures/}}
\begin{document}

\begin{frontmatter}

@@

%% Group authors per affiliation:
% \author{Matz Andreas Haugen\fnref{myfootnote}}
% \address{Enerhauggata 1, 0651 Oslo, Norway}
% \fntext[myfootnote]{Since 1880.}

%% or include affiliations in footnotes:
\author[mymainaddress, mysecondaryaddress]{Matz Andreas Haugen\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
% \ead[url]{www.matzhaugen.com}
\ead{matzhaugen@gmail.com}

\author[mysecondaryaddress]{Dorothea Gilbert}

\address[mymainaddress]{Enerhauggata 1, 0651, Oslo, Norway}
\address[mysecondaryaddress]{Independent Researcher, Oslo, Norway}

\begin{abstract}
Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor. 
Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction. Crucially, this augementation must be accompanied by causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people. We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
\end{abstract}

\begin{keyword}
Gompertz \sep Coherence \sep Covid \sep Coronavirus \sep Network Analysis \sep Stochastic
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}

Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \citep{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}. 
A large body of research reveals how these models fit the recent mortality seen due to the Coronavirus epidemic \citep{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.

However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825,bajzer1997mathematical} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
\begin{equation}
\label{eq:GompertzODE}
\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
\end{equation} 
% Redundant definitions for full script
with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
\begin{equation}
\label{eq:gomp_solution}
Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}},
\end{equation}
where $Y_{0}=Y(t = 0)$ and $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.

@@

This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves. These observations are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics. What is causing such a discrepancy between these observations and the current theory of communicable diseases \citep{castro2020turning}?

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same timeframe. While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model. We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.

\subsection{Simple communicable disease models}
We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge. First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I\nonumber,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

@@

 Due to this linearization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions. And since $I$ is a function of $Y$, we can combine Eq. \ref{eq:linSIR} via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
\begin{subequations}
\label{eq:rypdal}
\begin{align}
\frac{dY}{dt} =& \gamma(t) Y(t) \label{eq:rypdalODE}\\
\gamma(t) =& \frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right] \label{eq:rypdalGamma},
\end{align}
\end{subequations}
where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$. This parametrization is the commonly used Richards growth curve \citep{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \citep{wu2020generalized}. Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \citep{petroni2020logistic}. Note that at $\theta=1$, the traditional logistic growth curve is obtained, while at $\theta\rightarrow \infty$ we recover the exponential (Malthusian) explosion.

@@

\subsection{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease models. Under a double-log transform of Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. (2020) \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

@@

In the beginning of an epidemic one can assume small values of D and thus the SIRD model follows logistic growth with
\begin{equation}
\frac{dD}{dt} = c D (1 - D/K),
\end{equation}
obtained with a second order expansion of the exponential term in Eq. \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$. Both the SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.

Without this approximation, the full 3-parameters in Eq. \ref{eq:SIRD_D} of the diseased group can show a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq. \ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
\begin{equation}
\label{eq:GompertzODE2param}
\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right].
\end{equation}
Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \citep{roda2020difficult}.

The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating instantaneously as shown by Mombach et al. (2002) \cite{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain. We are thus prompted to look for another model which can explain the observed mortality patterns.

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text. Although the Gompertz curve can also be obtained with a simple linear fit using Eq. \ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day moving average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

\section{An alternative theory for observed mortality}
\label{seq:alt}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.

@@

This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}, where $Z(t)$ is Gaussian. It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}. As argued by De Lauro et al. (2014) \cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation. This is seen if we recast our perturbation model in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz. 
\begin{equation}
\label{eq:microscopicOriginal}
dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{C}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
\end{equation}

@@

Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \citep{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \citep{zhang1994log}. 
Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation \ref{eq:microscopicOriginal} \citep{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \citep{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}. 
Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (Eq. \ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$. This implies that the central limit theorem applies to the log-domain.

\section{Unifying the SIR model and the Gompertz model}
The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,

@@

Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by Petroni et al. (2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq. \ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''. 
Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

\subsection{Unification through a modified SIR model}
It is also possible to reconcile the SIR model with the Gompertz curve. 
Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
\begin{equation}
\label{eq:reverseNetworkSIR}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(1-x_j) \quad \forall i.
\end{equation} 
This crucial change is based on the environmental stressor hypothesis rather than a communicable disease assumption.
Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
\begin{equation}
\sum_i a_{ij} = 1 \quad \forall j.
\end{equation} 
If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model infection term becomes 
\begin{equation}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
\end{equation}

@@

\section{Conclusion}
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network. In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.

>>>>

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\newcommand{\insertSmallPdfFig}[3]{
  \begin{figure}[h]
  \centering
  \includegraphics[width=8.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}
\newcommand{\insertPdfFig}[3]{
  \begin{figure}[h]
  \centering
  \includegraphics[width=12.4cm]{#1.pdf}
  \caption{#2}
  \label{fig:#1}
  \end{figure}
}
\graphicspath{{../figures/}}

@@

\author[1,*]{Matz Haugen}
\author[2]{Dorothea Gilbert}
\affil[1]{Abel X, Oslo, Norway}
\affil[2]{Independent Researcher, Oslo, Norway}

\affil[*]{matzhaugen@gmail.com}

%\keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor. 
Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction. Crucially, this augementation must be accompanied by causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people. We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
\end{abstract}
\begin{document}

\flushbottom
\maketitle
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
\thispagestyle{empty}

\section*{Introduction}

Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}. 
A large body of research reveals how these models fit the recent mortality seen due to the Coronavirus epidemic \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.

However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \cite{Gompertz1825,bajzer1997mathematical} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
\begin{equation}
\label{eq:GompertzODE}
\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
\end{equation} 
% Redundant definitions for full script
with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
\begin{equation}
\label{eq:gomp_solution}
Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}},
\end{equation}
where $Y_{0}=Y(t = 0)$ and $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.

@@

This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves. These observations are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics. What is causing such a discrepancy between these observations and the current theory of communicable diseases \cite{castro2020turning}?

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same timeframe. While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model. We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.

\subsection*{Simple communicable disease models}
We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \cite{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge. First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I\nonumber,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

@@

 Due to this linearization, the infection/recovery ratio, $\phi(t)$, will have to change as a function of time to accommodate for the boundary conditions. And since $I$ is a function of $Y$, we can combine Eq. \ref{eq:linSIR} via an instantaneous relative growth rate, $\gamma(t) = dY(t) /(Ydt) = \beta I / Y$, in turn parameterized by a scaling factor, $\theta$, representing the shape of the growth,
\begin{subequations}
\label{eq:rypdal}
\begin{align}
\frac{dY}{dt} =& \gamma(t) Y(t) \label{eq:rypdalODE}\\
\gamma(t) =& \frac{\gamma_{\infty}}{\theta}\left[1 - \left(\frac{Y}{Y_{\infty}}\right)^{\theta} \right] \label{eq:rypdalGamma},
\end{align}
\end{subequations}
where $\gamma_{\infty} = \gamma(t\rightarrow \infty)$. This parametrization is the commonly used Richards growth curve \cite{richards1959flexible}, also called $\theta$-logistic growth, and has been used by others \cite{wu2020generalized}. Although not immediately justified in the communicable disease theory, one could imagine that $\theta$ represents non-linear network behavior \cite{petroni2020logistic}. Note that at $\theta=1$, the traditional logistic growth curve is obtained, while at $\theta\rightarrow \infty$ we recover the exponential (Malthusian) explosion.

@@

\subsection*{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease models. Under a double-log transform of Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. (2020) \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

@@

In the beginning of an epidemic one can assume small values of D and thus the SIRD model follows logistic growth with
\begin{equation}
\frac{dD}{dt} = c D (1 - D/K),
\end{equation}
obtained with a second order expansion of the exponential term in Eq. \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$. Both the SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \cite{vattay2020forecasting}.

Without this approximation, the full 3-parameters in Eq. \ref{eq:SIRD_D} of the diseased group can show a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq. \ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
\begin{equation}
\label{eq:GompertzODE2param}
\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right].
\end{equation}
Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \cite{roda2020difficult}.

The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating instantaneously as shown by Mombach et al. (2002) \cite{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain. We are thus prompted to look for another model which can explain the observed mortality patterns.

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text. Although the Gompertz curve can also be obtained with a simple linear fit using Eq. \ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day moving average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}

\section*{An alternative theory for observed mortality}
\label{seq:alt}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.

@@

This perturbation model is called an Ornstein–Uhlenbeck process \cite{risken1996fokker}, where $Z(t)$ is Gaussian. It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \cite{akaike1970statistical}. As argued by De Lauro et al. (2014) \cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation. This is seen if we recast our perturbation model in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz. 
\begin{equation}
\label{eq:microscopicOriginal}
dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{C}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
\end{equation}

@@

Thus, a more parsimonious interpretation of the observations not reliant on a transmission network is that mortality was caused by a planetary perturbation, modeled as a random process, to which organisms gradually develop resistance at a geometric rate in the log-transformed domain \cite{boxenbaum2017hypotheses,neafsey1988gompertz}, which is the natural transformation for many processes in nature \cite{zhang1994log}. 
Under this model, the distribution of the abundance of $F(t)$ is log-normal, a result that can be obtained by directly solving the stochastic equation \ref{eq:microscopicOriginal} \cite{skiadas2010exact,petroni2020gompertz}, or from thermodynamic principles \cite{sitaram1984statistical,gunasekaran1982lon,chakrabarti1996non}. 
Intuitively, this is seen by noting that the solution to the perturbation model in the log-domain (Eq. \ref{eq:microscopic}) is Gaussian in the variable $Z(t)$, thus suggesting a log-normal distribution of $F(t)$. This implies that the central limit theorem applies to the log-domain.

\section*{Unifying the SIR model and the Gompertz model}
The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,

@@

Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \cite{zhang1994log}.
We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by Petroni et al. (2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq. \ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''. 
Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

\subsection*{Unification through a modified SIR model}
It is also possible to reconcile the SIR model with the Gompertz curve. 
Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
\begin{equation}
\label{eq:reverseNetworkSIR}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(1-x_j) \quad \forall i.
\end{equation} 
This crucial change is based on the environmental stressor hypothesis rather than a communicable disease assumption.
Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
\begin{equation}
\sum_i a_{ij} = 1 \quad \forall j.
\end{equation} 
If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model infection term becomes 
\begin{equation}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
\end{equation}

@@

\section*{Conclusion}
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network. In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.

We further show that the observed mortality across countries can be explained by a model where the biological system is stressed by a ubiquitous and simultaneous stressor eliciting a corresponding stress response through which gradual return to pre-epidemic conditions are mediated. The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders. From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from quantum physics \cite{molski2003coherent}.
